A Mechanised Library for Generic (Co)Programming

### 2024-03-12 paper/paper.tex
Examples

So far, our examples could be modelled exclusively using catamorphisms. Here,
we show examples that require recursive hylomorphisms and termination proofs.

@@

\begin{minted}{OCaml}
  let rec msort x = match x with
| [] -> x | _ :: l -> (match l with | [] -> x | _ :: _ ->
     let (l0, r) = splitL x [] [] in
     let cx = fun p -> match p with | Lbranch -> l0 | Rbranch -> r in
     mergeL (msort (cx Lbranch)) (msort (cx Rbranch)))
\end{minted}

@@

@@

@@

>>>>

So far, our examples could be modelled exclusively using catamorphisms. Here,
we show examples that require recursive hylomorphisms and termination proofs.
We complete the sorting algorithm examples by applying fusion optimisation.

\subsubsection{Binary Trees}

Both Mergesort and Quicksort are hylomorphisms whose recursive structure is
described by a binary tree. In particular, the container used in their
definition is isomorphic to \coq{TreeF (list nat) unit} and 
\coq{TreeF unit nat} respectively, where \coq{TreeF} is defined as follows:
\begin{coqcode}
Inductive ITreeF L N X := i_leaf (l : L) | i_node (n : N) (l r : X)
\end{coqcode}
The definition as a container uses the following shapes and positions:
\begin{coqcode}
Inductive Tshape L A := | Leaf (ELEM : L) | Node (ELEM : A).
Inductive Tpos := | Lbranch | Rbranch. 
\end{coqcode}
We define the tree container, \coq{TreeF}, in a straightforward way, by making
the positions of type \coq{Tpos} only valid in \coq{Node}.
We define a series of definitions for tree container constructors and 
destructors:
\begin{coqcode}
Definition a_out {L A X : Type} : App (TreeF L A) X ~> ITreeF L A X.
(* Converting TreeF to ITreeF *)
Notation a_leaf x := (MkCont (Leaf _ x) (@dom_leaf _ _ _ x)).
Notation a_node x l r := (MkCont (Node _ x)
  (fun p => match val p with | Lbranch => l | Rbranch => r end)).
\end{coqcode}

@@

Mergesort is a hylomorphism with the recursive structure defined by the
container \coq{(TreeF (list int) unit)}. 
The computationally expensive part of the Mergesort implementation is the
algebra, since this is where most of the comparisons happen. The recursive
coalgebra for Mergesort is defined as the following Coq functions:
\begin{coqcode}
Fixpoint splitL (x : list int) (accL accR : list int) :=
  match x with
  | nil => (accL, accR)
  | cons x xs => splitL xs accR (cons x accL)
  end.

Definition c_split : Coalg (TreeF (list int) unit) (list int) :=
|{ x ~> match x with
        | nil | cons _ nil => a_leaf x
        | _ => let (l, r) := splitL x nil nil in
               a_node tt l r
        end
 }|.
Defined.
\end{coqcode}
Here, the notation \coq{|{ x ~> E }|} defines the morphism \coq{fun x => E},
and uses a custom tactic to prove that it is respectful. The recursive
coalgebra \coq{c_split} takes a list, and it returns a leaf with the input
list, if its size is less or equal than 1. Otherwise, it applies the function
\coq{splitL}, and divides the list in two halves.

It is easy to prove that this coalgebra is recursive, by showing that
it respects the relation \coq{<} on the length of the lists:
\begin{coqcode}
Lemma split_fin : respects_relation c_split (@length int) lt.
  Proof. (* *) Qed.

Definition tsplit : RCoalg (TreeF (list int) unit) (list int)
  := mk_wf_coalg wf_lt split_fin.
\end{coqcode}

The algebra simply applies the \coq{mergeL} function in order to the sublists
in every node:
\begin{coqcode}
Fixpoint mergeL l1 l2 {struct l1}  : list int := (* *)

Definition merge : App (TreeF (list int) unit) (list int) ~> list int.
|{ x ~> match a_out x with | node _ l r => mergeL l r | leaf e => e end }|.
Defined.
\end{coqcode}

Putting everything together, Mergesort is defined as the following
hylomorphism:
\begin{coqcode}
Definition msort := hylo merge tsplit.
\end{coqcode}
Note, however, that we could have composed the catamorphism and anamorphism,
and used the deforestation result to optimise the intermediate binary tree
away:
\begin{coqcode}
Definition msort : Ext (cata merge \o rana tsplit).
calculate. rewrite (deforest l_in_out). simpl. reflexivity. Defined.
\end{coqcode}
In both cases, the result is the same, and it extracts to the usual divide and
conquer definition of Mergesort in OCaml:
\begin{ocamlcode}
let rec msort x = match x with
| [] -> x | _ :: l -> (match l with | [] -> x | _ :: _ ->
     let (l0, r) = splitL x [] [] in
     let cx = fun p -> match p with | Lbranch -> l0 | Rbranch -> r in
     mergeL (msort (cx Lbranch)) (msort (cx Rbranch)))
\end{ocamlcode}

@@

The container for the Quicksort hylomorphism is \coq{TreeF unit int}, with
the following algebra and coalgebra.
\begin{coqcode}
  Definition merge : App (TreeF unit int) (list int) ~> list int.
|{ x : (App (TreeF unit int) (list int)) ~> (
           match x with
           | MkCont sx kx =>
               match sx return (Container.Pos sx -> _) -> _ with
               | Leaf _ _ => fun _ => nil
               | Node _ h => fun k => List.app (k (posL h)) (h :: k (posR h))
               end kx
           end
  )}|.
Defined.

Definition c_split : Coalg (TreeF unit int) (list int).
|{ x ~> match x with
        | nil => a_leaf tt
        | cons h t => let (l, r) := List.partition (fun x => x <=? h) t in
                      a_node h l r
        end}|.
Defined.
\end{coqcode}
As with the Mergesort case, we prove that the coalgebra is recursive by showing
that it respects the ``less-than'' relation on the length of the lists.
We then define, and extract, the hylomorphism \coq{hylo merge r_split}, which
extracts to the following OCaml code:
\begin{ocamlcode}
let rec qsort = function
| [] -> [] | h :: t ->
  let (l, r) = partition (fun x0 -> leb x0 h) t in
  let x0 = fun e -> qsort (match e with | Lbranch -> l | Rbranch -> r) in 
  app (x0 Lbranch) (h :: (x0 Rbranch))
\end{ocamlcode}
Note that Coq's code extraction is unable to inline \coq{x0}, but the resulting
code is similar to a hand-written \coq{qsort}.

@@

Suppose that we map a function to the result of sorting the list. We can
use our framework to fuse both computations. In particular, consider
the following definition:
\begin{coqcode}
Definition qsort_times_two : Ext (Lmap times_two \o hylo merge tsplit).
  calculate.  apply hylo_fusion_l. 
  (* calculate the fused stage *)
Defined.
\end{coqcode}
Here, \coq{Lmap times_two} is a list map function defined as a hylomorphism,
and \coq{times_two} multiplies every element of the list by two.
After applying hylo fusion and the necessary rewrites, the hylomorphism that we
extract is 
\coq{hylo (merge \o natural times_two) tsplit}. Here, \coq{natural} defines
a natural transformation between container extensions by applying a function
to the shapes that preserves the validity of the positions. In this case,
the function is \coq{times_two} that multiplies every pivot in the Quicksort
tree by two. The extracted OCaml code is a single recursive traversal:
\begin{ocamlcode}
let rec qsort_times_two = function
| [] -> []
| h :: t ->
  let (l, r) = partition (fun x0 -> leb x0 h) t in
  let x0 = fun p -> qsort_times_two (match p with
                                     | Lbranch -> l
                                     | Rbranch -> r) in
  app (x0 Lbranch) ((mul (Uint63.of_int (2)) h) :: (x0 Rbranch))
\end{ocamlcode}

@@

To close this section, we show how our framework can represent more advanced
recursion schemes, such as \emph{dynamorphisms} for dynamic programming, by
using their encoding as a hylomorphism. We use the knapsack example
from~\cite{HinzeWG15}. First, we define memoisation tables in terms
of containers. 
\begin{coqcode}
Definition MemoShape : Type := A * Sg.
Definition MemoPos := Pg.
Instance Memo : Cont MemoShape MemoPos := { valid := valid \o pair (snd \o fst) snd }.

Definition Table := LFix Memo.
\end{coqcode}
Memoisation tables are the least fixed point of the container defined by shapes
\coq{A * Sg} and positions \coq{Pg}, given a a container \coq{G : Cont Sg Pg}.
We define a function to insert elements into the memoisation tables:
\begin{coqcode}
  Definition Cons : A * App G Table ~> App Memo Table := (* *)
\end{coqcode}
And two functions to inspect the head of a memoisation table, and remove an
element from it:
\begin{coqcode}
Definition headT : Table ~> A := (* *)
Definition tailT : Table ~> App G Table := (* *)
\end{coqcode}
Using these definitions, a dynamorphism is defined as follows:
\begin{coqcode}
Definition dyna (a : App G Table ~> A) (c : RCoalg G B) : B ~> A
:= headT \o hylo (l_in \o Cons \o pair a id) c.
\end{coqcode}
Note how, instead of an algebra \coq{App G A ~> A}, the algebra takes a
memoisation table. The definition of the algebra can use this table to lookup
elements, instead of triggering a further recursive call. Elements are inserted
into the memoisation table by the use of \coq{Cons} to the result of applying
the algebra. The algebra for the dynamorphism looks up the previously computed
elements to produce the result, thus saving the corresponding recursive calls:
\begin{coqcode}
Fixpoint memo_knap table wvs :=
  match wvs with | nil => nil | h :: t =>
      match lookupT (Nat.pred (fst h)) table with
      | Some u => (u + snd h)%sint63 :: memo_knapsack table t
      | None => memo_knapsack table t
      end
  end.

Definition knapsack_alg (wvs : list (nat * int))
  (x : App NatF (Table NatF int)) : int :=
  match x with | MkCont sx kx => match sx with
  | inl tt => fun kx => 0%sint63
  | inr tt => fun kx => let tbl := kx next_elem in max_int 0 (memo_knap tbl wvs)
  end kx end.
Definition knapsackA wvs : App NatF (Table NatF int) ~> int := 
  (* [knapsack_alg wvs] as a respecful morphism *)
\end{coqcode}
The hylomorphism for knapsack is as follows, where \coq{out_nat} is the
recursive coalgebra for \coq{nat}.
\begin{coqcode}
Example knapsack wvs : Ext (dyna (knapsackA wvs) out_nat).
\end{coqcode}
Coq's code extraction mechanism is unable to inline several definitions in
this case, but the extracted code can be trivially inlined to produce the
following:
\begin{ocamlcode}
let knapsack wvs x =
  let (y, _) =
    (let rec f x0 =
      if x0=0 then Uint63.of_int (0)
      else let fn := f (x0-1) in { lFix_out = { 
           shape = (max_int (Uint63.of_int (0)) (memo_knap fn wvs), sx);
           cont = fun _ -> fn } }
     in f x).lFix_out.shape
  in y
\end{ocamlcode}
Note how the recursive calls of \ocaml{f} build the memoisation table, and how
this memoisation table is used to compute the intermediate results in
\ocaml{memo_knap}, which is finally discarded to produce the final result.
